---
title: "Week 4 In-Class: Exploratory Data Analysis"
author: "Syeda Aiman Fatima"
format: pdf
editor: visual
embed-resources: true
---

```{r}
#| label: load packages
#| include: false
library(here)
library(tidyverse)
library(ggplot2)
library(tinytex)

```

```{r}
#Check final data variables
finaldata <- read.csv(here("Data", "finaldata.csv"), header = TRUE)
names(finaldata)
```

```{r}
#Check observations from Canada

finaldata %>%
dplyr::filter(country_name == "Canada")

#Check observations from Ecuador

finaldata %>%
dplyr::filter(country_name == "Ecuador")
```

```{r}
summary(finaldata)

#missing data
sapply(finaldata, function(x) sum(is.na(x)))

#Identify countries with 20 missing values

missing_summary <- finaldata %>%
    select(country_name, ISO, Year, 
           popdens, urban, male_edu, temp, 
           rainfall1000, Infant_Mortality, 
           Neonatal_Mortality, Under5_Mortality) %>%
    mutate(popdens_missing = is.na(popdens),
           urban_missing = is.na(urban),
           male_edu_missing = is.na(male_edu),
           temp_missing = is.na(temp),
           rainfall1000_missing = is.na(rainfall1000),
           Infant_Mortality_missing = is.na(Infant_Mortality),
           Neonatal_Mortality_missing = is.na(Neonatal_Mortality),
           Under5_Mortality_missing = is.na(Under5_Mortality)) %>%
    filter(popdens_missing | urban_missing | male_edu_missing | 
           temp_missing | rainfall1000_missing | 
           Infant_Mortality_missing | Neonatal_Mortality_missing | 
           Under5_Mortality_missing) %>%
    select(country_name, ISO, Year, 
           popdens_missing, urban_missing, 
           male_edu_missing, temp_missing, 
           rainfall1000_missing, Infant_Mortality_missing, 
           Neonatal_Mortality_missing, Under5_Mortality_missing)

```

Missing data identified for all the variables. 20 missing values for each of the following variables: popdens, urban, male_edu, temp, rainfall1000, infant mortality, neonatal mortality, under5 mortality. In comparison, the variable gdp1000 has 62 missing values and maternal mortality has 426 missing values.

On further investigation, Cote d'Ivoire is the country with 20 missing values for the following variables: popdens, urban, male_edu, temp, & rainfall1000 from year 2000 to 2019. Puerto Rico is missing 20 values for infant, neonatal & under5 mortality from years 2000 to 2019.

```{r}
#Visualize distributions of mortality rates 

ggplot(finaldata, aes(x=Maternal_Mortality)) + 
  geom_histogram(binwidth=50) + 
  theme_minimal() + 
  labs(title="Distribution of Maternal Mortality")

ggplot(finaldata, aes(x=Infant_Mortality)) + 
  geom_histogram(binwidth=10) + 
  theme_minimal() + 
  labs(title="Distribution of Infant Mortality")

ggplot(finaldata, aes(x=Neonatal_Mortality)) + 
  geom_histogram(binwidth=5) + 
  theme_minimal() + 
  labs(title="Distribution of Neonatal Mortality")

ggplot(finaldata, aes(x=Under5_Mortality)) + 
  geom_histogram(binwidth=5) + 
  theme_minimal() + 
  labs(title="Distribution of Under 5 Mortality")
```

```{r}
#Visualize mortality trends 2000-2019

#Maternal Mortality 

finaldata %>%
  ggplot(aes(x = Year, y = Maternal_Mortality, group = ISO)) +
  geom_line(color = "skyblue") +
  xlim(c(2000,2019)) +
  labs(y = "Maternal mortality", x = "Year") + 
  theme_bw()

#Infant Mortality 

finaldata %>%
  ggplot(aes(x = Year, y = Infant_Mortality, group = ISO)) +
  geom_line(color = "skyblue") +
  xlim(c(2000,2019)) +
  labs(y = "Infant mortality", x = "Year") + 
  theme_bw()

#Neonatal Mortality

finaldata %>%
  ggplot(aes(x = Year, y = Neonatal_Mortality, group = ISO)) +
  geom_line(color = "skyblue") +
  xlim(c(2000,2019)) +
  labs(y = "Neonatal mortality", x = "Year") + 
  theme_bw()

#Under 5 Mortality

finaldata %>%
  ggplot(aes(x = Year, y = Under5_Mortality, group = ISO)) +
  geom_line(color = "skyblue") +
  xlim(c(2000,2019)) +
  labs(y = "Under 5 mortality", x = "Year") + 
  theme_bw()


```

The infant mortality trend graph shows a steep peak in mortality from 2009 to 2010 followed by a steep decline from 2010 to 2011. No discernible pattern for neonatal mortality. Similar peak and decline noticed for under 5 mortality as the infant mortality graph.

```{r}
#Visualize mortality trends by armed conflict status

#Maternal Mortality

finaldata %>%
  ggplot(aes(x = Year, y = Maternal_Mortality, group = ISO)) +
  geom_line(aes(color = as.factor(armconf1)), alpha = 0.5) +
  xlim(c(2000,2019)) +
  scale_y_continuous(trans='log10') + 
  labs(y = "Maternal mortality", x = "Year", color = "Armed conflict") + 
  theme_bw()

#Infant Mortality

finaldata %>%
  ggplot(aes(x = Year, y = Infant_Mortality, group = ISO)) +
  geom_line(aes(color = as.factor(armconf1)), alpha = 0.5) +
  xlim(c(2000,2019)) +
  scale_y_continuous(trans='log10') + 
  labs(y = "Infant mortality", x = "Year", color = "Armed conflict") + 
  theme_bw()

#Under 5 Mortality

finaldata %>%
  ggplot(aes(x = Year, y = Under5_Mortality, group = ISO)) +
  geom_line(aes(color = as.factor(armconf1)), alpha = 0.5) +
  xlim(c(2000,2019)) +
  scale_y_continuous(trans='log10') + 
  labs(y = "Under 5 mortality", x = "Year", color = "Armed conflict") + 
  theme_bw()

#Neonatal Mortality

finaldata %>%
  ggplot(aes(x = Year, y = Neonatal_Mortality, group = ISO)) +
  geom_line(aes(color = as.factor(armconf1)), alpha = 0.5) +
  xlim(c(2000,2019)) +
  scale_y_continuous(trans='log10') + 
  labs(y = "Neonatal mortality", x = "Year", color = "Armed conflict") + 
  theme_bw()
```

Similar trends observed for all 4 mortality outcomes, the highest mortality rates are observed where armed conflict is present.
